Supplementary Appendix 3: Trends in black and white mortality rates before and after excluding Hispanics, 1999-2013

### 2018-05-11 Manuscript/Appendix-1-agestd-mort-rates.Rmd
update appendices

---
title: 'Supplementary Appendix: Trends in the contribution of major causes of death
  to the black-white life expectancy gap by US state'
output:
  word_document: default
  html_document: default
---

This is the supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state".

For an interactive version of all our results, please visit XXXX.

```{r load_libraries, echo=FALSE, warning=F, message=F}
#Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Sam Harper, PhD; Jay S Kaufman, PhD

#Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, #1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(gridExtra)
library(plotly)
library(forcats)
source("../Code/life_expectancy_functions.R")

@@

**Table S1: Cause of death groupings**

@@

**Figure S3: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**

@@

**Figure S4: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**

@@

**Figure S5: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**

@@

**Figure S6: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**

@@

**Figure S7: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**

@@

**Figure S8: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**

@@

**Figure S9: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**

@@

**Figure S10: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**

@@

**Figure S11: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**

@@

**Figure S12: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**

@@

**Figure S13: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**

@@

**Figure S14: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**

@@

**Figure S15: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in males younger than one, 1969-2013**

@@

**Figure S16: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in females younger than one, 1969-2013**

>>>>

---
title: 'Supplementary Appendix: Trends in the contribution of major causes of death
  to the black-white life expectancy gap by US state'
output: 
  word_document:
      reference_docx: reference.docx
---

This is supplementary appendix #1 for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state".

Corinne A Riddell, PhD; Kathryn T Morrison, PhD; Jay S Kaufman, PhD; Sam Harper, PhD

Address correspondence to: Corinne Riddell, PhD, Department of Epidemiology, Biostatistics & Occupational Health, McGill University, 1020 Pine Avenue West, Room 36B, Montreal, QC H3A 1A2, Canada. Email: <corinne.riddell@mail.mcgill.ca>

For an interactive version of all our results, please visit https://corinne-riddell.shinyapps.io/black-white-life-expectancy/.

```{r load_libraries, echo=FALSE, warning=F, message=F}
library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(reshape2)
library(grid)
library(gridExtra)
library(plotly)
library(forcats)
source("../Code/life_expectancy_functions.R")

@@

#####
**Table S1: Cause of death groupings**

@@

#####PAGE BREAK

**Figure S1: Age-standardized trends in CVD-related mortality in black and white men, United States, 1969-2013**

@@

#####PAGE BREAK

**Figure S2: Age-standardized trends in CVD-related mortality in black and white women, United States, 1969-2013**

@@

#####PAGE BREAK 
**Figure S3: Age-standardized trends in cancer-related mortality in black and white men, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S4: Age-standardized trends in cancer-related mortality in black and white women, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S5: Age-standardized trends in mortality related to non-communicable disease in black and white men, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S6: Age-standardized trends in mortality related to non-communicable disease in black and white women, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S7: Age-standardized trends in injury-related mortality in black and white men, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S8: Age-standardized trends in injury-related mortality in black and white women, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S9: Age-standardized trends in mortality related to communicable disease in black and white men, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S10: Age-standardized trends in mortality related to communicable disease in black and white women, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S11: Age-standardized trends in mortality from all other causes in black and white men, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S12: Age-standardized trends in mortality from all other causes in black and white women, United States, 1969-2013**

@@

#####PAGE BREAK
**Figure S13: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in males infants (<1 year old), 1969-2013**

@@

#####PAGE BREAK
**Figure S14: Smoothed state-level trends in the contribution of all other causes to the life expectancy gap vs. the national pattern in female infants (<1 year old), 1969-2013**

### 2018-05-11 Manuscript/Appendix-2-compare-COD-gender-within-state.Rmd
update appendices

---
title: "Supplementary Appendix 2: Trend in the difference between each state's contribution trend and the overall trend, by causes of death and gender, 1969-2013"
output: word_document
---

This is the second supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013."

>>>>

---
title: "Supplementary Appendix 2: Trend in the difference between each state's contribution trend and the overall trend, by causes of death and gender, 1969-2013"
output: 
  word_document:
      reference_docx: reference.docx
---

This is the second supplementary appendix #2 for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013."

### 2018-05-11 Manuscript/Appendix-3-Explore-Hispanic-origin.Rmd
update appendices

---
title: "Supplementary Appendix 3: Trends in black and white mortality rates before and after excluding Hispanics, 1999-2013"
output: github_document
---

This is the third supplementary appendix for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013."

@@

In our main analyses, we included Hispanics as part of the black and white racial groups. To investigate the impact of including Hispanics on mortality, we inspected mortality trends for five key states with large Hispanic immigrant populations (California, New York, Florida, Texas, and Illinois). We extracted this mortality data using (CDC WONDER's)[https://wonder.cdc.gov/] compressed mortality file. Hispanic ethnicity is only available beginning in 1999, so we consider the time period from 1999 to 2013. We extract age-adjusted mortality rates related to the circulatory system ("CVD", ICD-10 codes I00-I99), neoplasms ("cancers", ICD-10 codes C00-D48) and external causes ("injuries", ICD-10 codes V01-Y89) by year, state, race, and gender, for (a) non-Hispanics and (b) regardless of Hispanic ethinicity.

**Trends in cancer mortality for all whites and blacks vs. non-Hispanic whites and blacks, 1999-2013**
```{r load-cancer-data, echo=F, fig.width = 8, fig.height=9.5}
library(tidyverse)
nh.cancer <- read.table("../Data/hispanic/NH Black and White Neoplasms.txt", header = T, nrows = 360, sep="\t")

>>>>

---
title: "Supplementary Appendix 3: Trends in black and white mortality rates before and after excluding Hispanics, 1999-2013"
output: 
  word_document:
      reference_docx: reference.docx
---

This is the third supplementary appendix #3 for the paper, "Trends in the contribution of major causes of death to the black-white life expectancy gap by US state. A retrospective study of all deaths between 1969 and 2013."

@@

#####
In our main analyses, we included Hispanics as part of the black and white racial groups. To investigate the impact of including Hispanics on mortality, we inspected mortality trends for five key states with large Hispanic immigrant populations (California, New York, Florida, Texas, and Illinois). We extracted this mortality data using (CDC WONDER's)[https://wonder.cdc.gov/] compressed mortality file. Hispanic ethnicity is only available beginning in 1999, so we consider the time period from 1999 to 2013. We extract age-adjusted mortality rates related to the circulatory system ("CVD", ICD-10 codes I00-I99), neoplasms ("cancers", ICD-10 codes C00-D48) and external causes ("injuries", ICD-10 codes V01-Y89) by year, state, race, and gender, for (a) non-Hispanics and (b) regardless of Hispanic ethinicity.

#####
**Trends in cancer mortality for all whites and blacks vs. non-Hispanic whites and blacks, 1999-2013**
```{r load-cancer-data, echo=F, fig.width = 8, fig.height=9.5, warning=F, message=F}
library(tidyverse)
nh.cancer <- read.table("../Data/hispanic/NH Black and White Neoplasms.txt", header = T, nrows = 360, sep="\t")
